---
title: "GROUP-8 PROJECT ANALYSIS"
author: "GROUP-8"
date: "2023-04-30"
output: pdf_document
---

## **INTRODUCTION:**

We have selected the brain stroke dataset, which is available on Kaggle and contains information on patients who have experienced a stroke. We aim to investigate the potential links between work status, hypertension, and glucose levels and the incidence of brain stroke. Additionally, we also want to examine whether age and average glucose level could potentially act as mediators in this relationship and provide insights on how these factors may be interrelated. The objective of our study is to identify any important factors that may contribute to the likelihood of brain stroke occurrence in the population.

## **Importing and viewing file:**

```{r}
library(readxl)
brain_stroke <- read_excel("C:/Users/madhu/Downloads/brain_stroke.xlsx")
View(brain_stroke)
```

```{r}
# viewing number of rows and columns in the data set
dim(brain_stroke)
```

```{r}
head(brain_stroke)
```

```{r}
#viewing data Categories of variables
str(brain_stroke)
```

## **Summary statistics:**

```{r}
summary(brain_stroke)
```

## **Exploratory data analysis:**

#### **Plotting histograms to check normality of the data**

1)  For average glucose levels

```{r}
hist(brain_stroke$avg_glucose_level)
```

2)  For age

```{r}
hist(brain_stroke$age)
```

3)  For BMI

```{r}
hist(brain_stroke$bmi)
```

**Q-Q PLOTS FOR NORMALITY**

```{r}
qqnorm(brain_stroke$avg_glucose_level)
```

```{r}
qqnorm(brain_stroke$age)

```

```{r}
qqnorm(brain_stroke$bmi)
```

#### **Plotting bar plots**

1)  For hypertension

```{r}
# One way table to summarize hypertension data
table(brain_stroke$hypertension)
```

```{r}
# plotting bar plot for hypertension
barplot(table(brain_stroke$hypertension))
```

2)  For work_type

```{r}
# One way table to summarize work_type data
table(brain_stroke$work_type)
```

```{r}
# plotting bar plot for work_type
barplot(table(brain_stroke$work_type))
```

3)  For stroke

```{r}
# One way table to summarize stroke data
table(brain_stroke$stroke)
```

```{r}
# plotting bar plot for stroke
barplot(table(brain_stroke$stroke))
```

#### **Plotting box plot to check for outliers in data:**

1)  For age

```{r}
boxplot(brain_stroke$age~brain_stroke$stroke)

```

2)  For average glucose levels

```{r}
boxplot(brain_stroke$avg_glucose_level~brain_stroke$stroke)
```

## **Importing and viewing the new data set after categorizing and coding variables:**

```{r}
library(readxl)
brain_stroke1 <- read_excel("C:/Users/madhu/Downloads/brain_stroke1.xlsx")
View(brain_stroke1)
```

## **Sampling:**

```{r}
#installing package sampling to perform stratified random sampling
#install.packages("sampling")
library("sampling")
```

#### Performing random stratification of sample considering gender, heart_disease, and bmi as confounding variables.

```{r}
set.seed(1)
strat.samp <- strata(brain_stroke1, stratanames = 
c("gender", "heart_disease", "bmi_status"), size = rmultinom(30, 30, 30),
method = "srswr")
strat.samp
```

```{r}
# Taking a stratified random sample as bs.sample
num_samples=5
for (i in 1:num_samples) {bs.samp <- brain_stroke1[strat.samp$ID_unit, ]}
bs.samp
```

#### Simple random sampling:

```{r}
#set.seed(1)
#sample.size = sample(1:nrow(brain_stroke1), 500, replace = FALSE)
#bs.samp = brain_stroke1[sample.size, ]
#bs.samp
```

## Correlation test:

```{r}
cor1 = cor.test(bs.samp$stroke, bs.samp$age, method = "spearman")
cor1
cor2 = cor.test(bs.samp$stroke, bs.samp$work, method = "spearman")
cor2
cor3 = cor.test(bs.samp$stroke, bs.samp$blood_glucose, method = "spearman")
cor3
cor4 = cor.test(bs.samp$stroke, bs.samp$blood_glucose2, method = "spearman")
cor4
```

[Interpretation:]{.underline}

1.  For the correlation between **`bs.samp$stroke`** and **`bs.samp$age`**, the Spearman's rank correlation coefficient was estimated as 0.259 and the p-value was estimated as 9.315e-10. This suggests that there is a strong positive correlation between stroke and age in the sample.
2.  For the correlation between **`bs.samp$stroke`** and **`bs.samp$work`**, the Spearman's rank correlation coefficient was estimated as 0.0831 and the p-value was estimated as 0.05361. This suggests that there may be a weak positive correlation between stroke and work in the sample, but this correlation is not statistically significant at the conventional 5% level.
3.  For the correlation between **`bs.samp$stroke`** and **`bs.samp$blood_glucose`**, the Spearman's rank correlation coefficient was estimated as 0.306 and the p-value was estimated as 3.72e-13. This suggests that there is a strong positive correlation between stroke and blood glucose in the sample.
4.  For the correlation between **`bs.samp$stroke`** and **`bs.samp$blood_glucose2`**, the Spearman's rank correlation coefficient was estimated as 0.292 and the p-value was estimated as 4.661e-12. This suggests that there is a strong positive correlation between stroke and blood glucose level measured after a meal in the sample.
5.  Overall, these results suggest that age, blood glucose, and blood glucose level after a meal are strongly positively correlated with stroke, while work may have a weaker positive association with stroke that is not statistically significant in this sample. However, these results are based on a single sample, and the associations may differ in other populations.

## **Unpaired two sample t-test:**

un-paired two Sample t-test for comparing the mean stroke status of data elements with two categories i.e., average glucose levels

```{r}
ndiabetic = subset(bs.samp, blood_glucose == 0)
diabetic = subset(bs.samp, blood_glucose == 1)
t.test(ndiabetic$stroke, diabetic$stroke, na.rm = TRUE,
       conf.level = 0.95, alternative = "two.sided")
```

[Interpretation:]{.underline}

The Welch two-sample t-test was conducted to determine if there was a significant difference in the mean values of the variable 'stroke' between two groups, 'ndiabetic' and 'diabetic'. The test statistic (t) was calculated to be -4.8835 with a degrees of freedom (df) of 99.291, and a very small p-value of 3.993e-06. This indicates strong evidence against the null hypothesis of no difference between the means of the two groups. Therefore, we reject the null hypothesis and conclude that there is a significant difference between the mean values of 'stroke' for the two groups.

The alternative hypothesis suggests that the true difference in means is not equal to 0. The 95% confidence interval of the difference between means is (-0.3389517, -0.1430964), which does not include 0. This indicates that the difference between the means is statistically significant.

The sample estimates for the mean value of 'stroke' for group 'ndiabetic' and 'diabetic' are 0.05567929 and 0.29670330, respectively. This suggests that patients with diabetes have a higher mean value for 'stroke' compared to those without diabetes.

## **Chi-square test:**

1\. For age groups:

```{r}
# Creating a two-way contingency variable in R
table = table(brain_stroke1$stroke, brain_stroke1$age_groups)
table
# Chi-square test
chisq.test(table)
```

[Interpretation:]{.underline}

The result of the Pearson's Chi-squared test indicates that there is a significant association between the variables in the contingency table.

The test statistic (X-squared) is 314, which is large and suggests that the observed counts in the table are different from what would be expected if there was no association between the variables.

The degrees of freedom (df) are 5, which corresponds to the number of categories in one variable minus 1, times the number of categories in the other variable minus 1.

The p-value is less than 2.2e-16, which is essentially zero. This indicates that the probability of observing such an extreme test statistic by chance, assuming no association between the variables, is extremely low.

Therefore, we can reject the null hypothesis of no association and conclude that there is a significant association between the variables in the contingency table.

2.  For work:

    ```{r}
    # Creating a two-way contingency variable in R
    table = table(brain_stroke1$stroke, brain_stroke1$work_type)
    table
    # Chi-square test
    chisq.test(table)
    ```

    Interpretation:

    The Pearson's chi-squared test shows a statistically significant result with a chi-squared value of 47.832 and 3 degrees of freedom (df), which corresponds to a very small p-value of 2.312e-10. This indicates that there is strong evidence to reject the null hypothesis that the observed frequencies in the contingency table are equal to the expected frequencies, and that there is a significant association between the two categorical variables being analyzed.

3.  For average blood glucose levels in 2 categories i.e., non-diabetic and diabetic:

    ```{r}
    # Creating a two-way contingency variable in R
    table = table(brain_stroke1$stroke, brain_stroke1$blood_glucose)
    table
    # Chi-square test
    chisq.test(table)
    ```

    Interpretation:

    The result of the Pearson's Chi-squared test with Yates' continuity correction indicates that there is a significant association between the variables in the contingency table.

    The test statistic (X-squared) is 63.617, which is large and suggests that the observed counts in the table are different from what would be expected if there was no association between the variables.

    The degrees of freedom (df) are 1, which corresponds to the number of categories in one variable minus 1, times the number of categories in the other variable minus 1, minus 1 for the correction.

    The p-value is 1.511e-15, which is essentially zero. This indicates that the probability of observing such an extreme test statistic by chance, assuming no association between the variables, is extremely low.

    Therefore, we can reject the null hypothesis of no association and conclude that there is a significant association between the variables in the contingency table, even after applying Yates' continuity correction to adjust for the small expected cell frequencies.

4.  For average blood glucose in 4 categories i.e., low, normal, pre-diabetic and diabetic:

```{r}
#Creating a two-way contingency variable in R
table = table(brain_stroke1$stroke, brain_stroke1$blood_glucose2)
table
# Chi-square test
chisq.test(table)
```

Interpretation:

The result of the Pearson's Chi-squared test indicates that there is a significant association between the variables in the contingency table.

The test statistic (X-squared) is 86.324, which is large and suggests that the observed counts in the table are different from what would be expected if there was no association between the variables.

The degrees of freedom (df) are 3, which corresponds to the number of categories in one variable minus 1, times the number of categories in the other variable minus 1.

The p-value is less than 2.2e-16, which is essentially zero. This indicates that the probability of observing such an extreme test statistic by chance, assuming no association between the variables, is extremely low.

Therefore, we can reject the null hypothesis of no association and conclude that there is a significant association between the variables in the contingency table.

## **Logistic Regression:**

1.  For age groups

```{r}
model <- glm(stroke ~ age_groups, data = bs.samp,
             family = "binomial"(link = "logit"))
summary(model)
```

[Interpretation:]{.underline}

The coefficients table shows that age_groups is a significant predictor of the likelihood of having a stroke (z = 4.459, p \< 0.001), with an estimated coefficient of 1.2148. The negative intercept coefficient (-7.5720) indicates that the probability of having a stroke decreases as age increases.

The deviance residuals describe the difference between the predicted and observed values for the dependent variable. The values range from -0.6352 to 2.3567, with higher values indicating a larger difference between predicted and observed values.

The AIC (Akaike Information Criterion) is a measure of the model's goodness-of-fit, with lower values indicating a better fit. The AIC value for this model is 315.91.

In conclusion, the logistic regression model suggests that age_groups is a significant predictor of the likelihood of having a stroke, with older age groups having a higher probability of having a stroke.

2.  For work type

    ```{r}
    model <- glm(stroke ~ work, data = bs.samp,
                 family = "binomial"(link = "logit"))
    summary(model)
    ```

    [Interpretation:]{.underline}

    The coefficients table shows that work is a significant predictor of the likelihood of having a stroke (z = 2.501, p = 0.0124), with an estimated coefficient of 0.4715. The positive coefficient indicates that being employed is associated with a higher probability of having a stroke, compared to being unemployed or retired.

    The deviance residuals describe the difference between the predicted and observed values for the dependent variable. The values range from -0.5645 to 2.3512, with higher values indicating a larger difference between predicted and observed values.

    The AIC (Akaike Information Criterion) is a measure of the model's goodness-of-fit, with lower values indicating a better fit. The AIC value for this model is 339.12.

    In conclusion, the logistic regression model suggests that work status is a significant predictor of the likelihood of having a stroke, with employed individuals having a higher probability of having a stroke compared to unemployed or retired individuals. However, the effect size of work on stroke is smaller compared to age, as indicated by the smaller coefficient and smaller difference between null and residual deviances.

3.  For blood glucose: non-diabetic and diabetic

    ```{r}
    model <- glm(stroke ~ blood_glucose, data = bs.samp,
                 family = "binomial"(link = "logit"))
    summary(model)
    ```

    [Interpretation:]{.underline}

    The coefficients table shows that blood glucose is a significant predictor of the likelihood of having a stroke (z = 6.384, p \< 0.001), with an estimated coefficient of 1.9678. The positive coefficient indicates that higher blood glucose levels are associated with a higher probability of having a stroke.

    The deviance residuals describe the difference between the predicted and observed values for the dependent variable. The values range from -0.8390 to 2.4034, with higher values indicating a larger difference between predicted and observed values.

    The AIC (Akaike Information Criterion) is a measure of the model's goodness-of-fit, with lower values indicating a better fit. The AIC value for this model is 307.65.

    In conclusion, the logistic regression model suggests that blood glucose levels are a significant predictor of the likelihood of having a stroke, with higher levels of blood glucose being associated with a higher probability of having a stroke. The effect size of blood glucose on stroke is larger than that of work status or age, as indicated by the larger coefficient and the larger difference between null and residual deviance.

4.  For blood glucose: low, normal, pre-diabetic, diabetic

    ```{r}
    model <- glm(stroke ~ blood_glucose2, data = bs.samp,
                 family = "binomial"(link = "logit"))
    summary(model)
    ```

    [Interpretation:]{.underline}

    The model indicates that for a one unit increase in blood_glucose2, the odds of having a stroke increase by a factor of exp(1.0691) = 2.91, holding all other predictors constant. The intercept of -4.0523 represents the log odds of having a stroke when blood_glucose2 is equal to zero.

    The p-value for the blood_glucose2 coefficient is less than 0.001, indicating that it is statistically significant at a 0.05 level of significance. The null deviance is the deviance of the model with only the intercept, and the residual deviance is the deviance of the model with the predictor variable. The difference between the null and residual deviances suggests that the blood_glucose2 variable improves the model fit.

    The AIC value of 298 is lower than the AIC values for the other models, indicating that this model has the best balance between model fit and complexity among the four models. Overall, the model suggests that blood glucose level is a significant predictor of stroke risk.

Looking at the output of the for the logistic regression models with blood_glucose and blood_glucose2, we can see that the AIC for the blood_glucose model is 307.65, while the AIC for the blood_glucose2 model is 298. Therefore, the blood_glucose2 model is considered better because it has a lower AIC.

Therefore, including blood_glucose2 for combined and interaction effect.

2.  For combined effect

```{r}
model <- glm(stroke ~ age_groups + work + blood_glucose2,
             data = bs.samp, family = "binomial"(link = "logit"))
summary(model)
```

[Interpretation:]{.underline}

The coefficients show that age and blood glucose levels are significant predictors of stroke occurrence, while work is not as significant as the prior two. Specifically, for every one-unit increase in age groups, the odds of stroke increase by a factor of 2.5. For every one-unit increase in blood glucose levels, the odds of stroke increase by a factor of 2.4.

[**Final Conclusion:**]{.underline} The current model has the least AIC score i.e., 290 compared to previous models. Therefore, the model with combined effect of work, age groups, and blood glucose levels is the best model so far.

3.  For interaction effect

    ```{r}
    model <- glm(stroke ~ age_groups * work * blood_glucose2,
                 data = bs.samp, family = "binomial"(link = "logit"))
    summary(model)
    ```

[Interpretation:]{.underline}

Intercept: -12.7519, indicating the log odds of stroke for the reference category when all other predictors are at zero.

Age groups: The coefficient for age groups is 1.7202, which is not statistically significant (p = 0.560), suggesting that the effect of age groups on stroke occurrence is not significant after accounting for the other predictors in the model.

Work: The coefficient for work is -1.7135, which is also not statistically significant (p = 0.785), suggesting that work may not be a significant predictor of stroke occurrence in this model.

Blood glucose levels: The coefficient for blood glucose levels is -1.2923, which is not statistically significant (p = 0.841), indicating that blood glucose levels may not be a significant predictor of stroke occurrence in this model.

Age groups:work: The interaction term between age groups and work is 0.4971, which is not statistically significant (p = 0.708), suggesting that the effect of the interaction between age groups and work on stroke occurrence is not significant after accounting for the other predictors in the model.

Age groups:blood_glucose2: The interaction term between age groups and blood glucose levels is 0.6941, which is not statistically significant (p = 0.617), suggesting that the effect of the interaction between age groups and blood glucose levels on stroke occurrence is not significant after accounting for the other predictors in the model.

Work:blood_glucose2: The interaction term between work and blood glucose levels is 3.2156, which is not statistically significant (p = 0.279), indicating that the effect of the interaction between work and blood glucose levels on stroke occurrence may not be significant in this model.

Age groups:work:blood_glucose2: The interaction term between age groups, work, and blood glucose levels is -0.7973, which is not statistically significant (p = 0.209), suggesting that the effect of the interaction between age groups, work, and blood glucose levels on stroke occurrence is not significant after accounting for the other predictors in the model.

Overall, the model does not provide strong evidence that any of the included variables or interaction terms are significant predictors of stroke occurrence after accounting for the other variables in the model. However, it is important to note that this model may not be the best-fitting or most appropriate model for this data set, and further analysis may be needed to determine the most appropriate predictors of stroke occurrence.

## Studying interaction effect of hypertension and blood glucose on work type's association with stroke:

```{r}
model <- glm(stroke ~ work * hypertension * blood_glucose2,
             data = bs.samp, family = "binomial"(link = "logit"))
summary(model)
```

[Interpretation:]{.underline}

The intercept coefficient is -7.7361, indicating that the log odds of stroke when all the independent variables are at 0 is -7.7361.

The coefficient for work is 1.7415, indicating that the log odds of stroke increase by 1.7415 when work increases by 1 unit, holding all other variables constant.

The coefficient for hypertension is 2.4800, indicating that the log odds of stroke increase by 2.4800 when hypertension is present, holding all other variables constant.

The coefficient for blood glucose is 2.8734, indicating that the log odds of stroke increase by 2.8734 when blood glucose increases by 1 unit, holding all other variables constant.

The coefficient for work:hypertension is -0.8358, indicating that the effect of work on stroke is moderated by hypertension such that the log odds of stroke decrease by 0.8358 when work and hypertension increase by 1 unit, holding blood glucose constant.

The coefficient for work:blood_glucose2 is -0.8802, indicating that the effect of work on stroke is moderated by blood glucose such that the log odds of stroke decrease by 0.8802 when work and blood glucose increase by 1 unit, holding hypertension constant.

The coefficient for hypertension:blood_glucose2 is -1.4210, indicating that the effect of hypertension on stroke is moderated by blood glucose such that the log odds of stroke decrease by 1.4210 when hypertension and blood glucose increase by 1 unit, holding work constant.

The coefficient for work:hypertension:blood_glucose2 is 0.6376, indicating that the effect of work on stroke is moderated by both hypertension and blood glucose such that the log odds of stroke increase by 0.6376 when work, hypertension, and blood glucose increase by 1 unit.

```{r}
library(caret)

# Split the data into training and testing sets (70% training, 30% testing)
set.seed(123)
trainIndex <- createDataPartition(bs.samp$stroke, p = 0.7, list = FALSE)
train_set <- bs.samp[trainIndex, ]
test_set <- bs.samp[-trainIndex, ]


# Model : For all variables
model4 <- glm(stroke ~ age_groups + work + blood_glucose2, data = bs.samp, family = "binomial"(link = "logit"))
pred_probs4 <- predict(model4, newdata = test_set, type = "response")
pred_labels4 <- ifelse(pred_probs4== "Yes", 1, 0)
confusionMatrix4 <- confusionMatrix(factor(pred_labels4), factor(test_set$stroke))
print(confusionMatrix4)
```
